Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-08-15 numerics_paper/version2/numerical_chj.tex

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and Rusanov suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical} and \cite{zaide2012numerical}.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.  Another approach to avoid
shallow water carbuncles has been proposed in \cite{navas2019improved}, where it is
studied together with the slowly-moving shock anomaly and in the presence of varying bathymetry.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe.
%based on a local measure of the entropy residual, 
%as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
{\color{red} Since Rusanov's solver does not suffer from carbuncle instabilities,
  our aim is to use it near shocks without degrading the good accuracy of Roe's solver. 
  To do this, we need to find a blending criteria that can robustly and sharply locate shocks.
  The entropy residual has proven to be a successful tool for this task; see for instance
  \cite{guermond2011entropy,guermond2018second,guermond2018well}.
  By doing this, our blended solver induces dissipation of entropy.
  Entropy dissipative Riemann solvers have been used to produce carbuncle free solvers for the Euler equations;
  see for instance \cite{ismail2009affordable,ismail2009proposed}.
  However, to the best of our knowledge, this idea has not been used before to supress carbuncle instabilities for the
  shallow water equations. 
}
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.

@@

We model only the flow on the upstream side of the cylinder, since
our interest is in the resolution of the bow shock.  We take the domain
$[0,40]\times[0,100]$ with a cylinder of radius 20 centered at $(40,50)$.
Reflecting boundary conditions are imposed at the surface of the cylinder,
along with outflow conditions along the rest of the right edge of the domain.
The depth and velocity are set initially and (for all time) at the other
boundaries to $h_0=1$, $u_0=5$, for a Froude number of 5.
We use a $160 \times 400$ uniform Cartesian grid.
%
In Figure \ref{fig:steady_outflow}, we show results at $t=80$, after the flow has reached
a steady state.  With Roe's solver, negative values (of the water height) are generated at an early
time, leading to failure of the solver.  Therefore, we show results only for
Rusanov's, HLLEMCC by \cite{kemm2014note}, and the blended solvers.
{\color{red}
  These three methods give carbuncle free results.
  It is worth noting that the solution with Rusanov's solver is the most dissipated solution. 
  Let 
  \begin{align}\label{res}
    \xi(t^{n+1}):=\frac{||h^{n+1}-h^n||_{\ell^2}}{||h^0||_{\ell^2}}
  \end{align}
  denote the residual based on the water height. In Figure \ref{fig:residuals_cylinder_2nd}, we plot $\xi(t)$ for the
  solution obtained with each Riemann solver. With HLLEMCC or the blended solver, the residual stagnates. 
  In Figure \ref{fig:residuals_cylinder_1st} we repeat the same experiment but using the first-order method \eqref{first-order_via_fluct} instead.
  We obtain similar results with Rusanov and HLLEMCC, but now the solution with the blended solver continues to converge
  at a rate similar to that of Rusanov.  This suggests that the lack of convergence for the 2nd-order method with the
  blended solver is due to the correction terms and not to the Riemann solver itself.
  In Figure \ref{fig:cylinder_blended}, we show a closeup of the front of the shock and show the grid we use.
  Note that the grid is cartesian and that the shock is resolved across a few
  cells. Therefore, the ability of the method to deliver
  carbuncle free solutions does not depend on the shock being aligned with a grid interface. 
}

>>>>

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and Rusanov suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical} and \cite{zaide2012numerical}.
{\color{OliveGreen}
  The slowly-moving shock anomaly is another important unphysical behavior that might be present 
  with some numerical schemes. We refer to \cite{zaide2012numerical} for a comprehensive 
  discussion about this phenomenon.
  In \cite{navas2019improved}, the authors focus on this problem when solving the shallow water equations. 
}

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.  Another approach to avoid
shallow water carbuncles has been proposed in \cite{navas2019improved}, where it is
studied together with the slowly-moving shock anomaly and in the presence of varying bathymetry.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe.
%based on a local measure of the entropy residual, 
%as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
{\color{red} Since Rusanov's solver does not suffer from carbuncle instabilities,
  our aim is to use it near shocks without degrading the good accuracy of Roe's solver. 
  To do this, we need to find a blending criteria that can robustly and sharply locate shocks.
  The entropy residual has proven to be a successful tool for this task; see for instance
  \cite{guermond2011entropy,guermond2018second,guermond2018well}.
  By doing this, our blended solver induces dissipation of entropy.
  Entropy dissipative Riemann solvers have been used to produce carbuncle free solvers for the Euler equations;
  see for instance \cite{ismail2009affordable,ismail2009proposed}.
  However, to the best of our knowledge, this idea has not been used before to supress carbuncle instabilities for the
  shallow water equations. 
}
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.  
{\color{OliveGreen}
  Based on our numerical experiments, using Rusanov's solver does not lead to the 
  slowly-moving shock anomaly. Our blended solver inherits the same property.
}

@@

We model only the flow on the upstream side of the cylinder, since
our interest is in the resolution of the bow shock.  We take the domain
$[0,40]\times[0,100]$ with a cylinder of radius 20 centered at $(40,50)$.
Reflecting boundary conditions are imposed at the surface of the cylinder,
along with outflow conditions along the rest of the right edge of the domain.
{\color{red} The depth and velocity are set initially and (for all time) at the other
  boundaries to $h_0=1$, $u_0=5$, $v_0=0$ for a Froude number of 5. Therefore, 
  the initial condition does not contain a shock, but due to the presence of the 
  cylinder a bow shock rapidly forms and travels to its eventual steady-state position.}
  We use a $160 \times 400$ uniform Cartesian grid.
%
In Figure \ref{fig:steady_outflow}, we show results at $t=80$, after the flow has reached
a steady state.  With Roe's solver, negative values (of the water height) are generated at an early
time, leading to failure of the solver.  Therefore, we show results only for
Rusanov's, HLLEMCC by \cite{kemm2014note}, and the blended solvers.
{\color{red}
  These three methods give carbuncle free results.
  It is worth noting that the solution with Rusanov's solver is the most dissipated solution. 
  Let 
  \begin{align}\label{res}
    \xi(t^{n+1}):=\frac{||h^{n+1}-h^n||_{\ell^2}}{||h^0||_{\ell^2}}
  \end{align}
  denote the residual based on the water height. In Figure \ref{fig:residuals_cylinder_2nd}, we plot $\xi(t)$ for the
  solution obtained with each Riemann solver. With HLLEMCC or the blended solver, the residual stagnates. 
  In Figure \ref{fig:residuals_cylinder_1st} we repeat the same experiment but using the first-order method \eqref{first-order_via_fluct} instead.
  We obtain similar results with Rusanov and HLLEMCC, but now the solution with the blended solver continues to converge
  at a rate similar to that of Rusanov.  This suggests that the lack of convergence for the 2nd-order method with the
  blended solver is due to the correction terms and not to the Riemann solver itself.
  In Figure \ref{fig:cylinder_blended}, we show a closeup of the front of the shock and show the grid we use.
  Note that the grid is cartesian and that the shock is resolved across a few
  cells. Therefore, the ability of the method to deliver
  carbuncle free solutions does not depend on the shock being aligned with a grid interface. 
}
